# Clustering and Classification

This exercise analyses data from the *Boston Dataset*, which is a dataset containing Housing Values of Boston, Massachusettes Suburbias.  

```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# access the MASS package
library(MASS)
# load the data
data("Boston")
```
The data has 506 observations and 14 variables  
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Exploring the dimensions of the data
dim(Boston)
```
and its data structure looks like this:
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Exploring the structure of the data
str(Boston)
```
  
#####The variables stand for:
* crim - per capita crime rate by town.
* zn - proportion of residential land zoned for lots over 25,000 sq.ft.
* indus - proportion of non-retail business acres per town.
* chas - Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).
* nox - nitrogen oxides concentration (parts per 10 million).
* rm - average number of rooms per dwelling.
* age - proportion of owner-occupied units built prior to 1940.
* dis - weighted mean of distances to five Boston employment centres.
* rad - index of accessibility to radial highways.
* tax - full-value property-tax rate per \$10,000.
* ptratio - pupil-teacher ratio by town.
* black - 1000(Bk - 0.63)^2^ where Bk is the proportion of blacks by town.
* lstat - lower status of the population (percent).
* medv - median value of owner-occupied homes in \$1000s.
  
#####The source of the date are  
Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. J. Environ. Economics and Management 5, 81–102.  
Belsley D.A., Kuh, E. and Welsch, R.E. (1980) Regression Diagnostics. Identifying Influential Data and Sources of Collinearity. New York: Wiley.  
  
  
#####Now, let's print a summary of the data:  
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Getting the summary of the data
summary(Boston)
```
#####and visualize the data by printing its correlation matrix  
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
library(corrplot, tidyverse)
# calculate the correlation matrix
cor_matrix<-cor(Boston)
# print the correlation matrix
corrplot(cor_matrix, type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
```

As we can see from the matrix, some variables have a strong positive (blue circles) or negative correlation (red circles). It appears, for example, that the area's property-tax rate (tax) correaltes heavily with access to radial highways (rad). In ther words, the closer the highways, the bigger the tax.  
  
The area's distance from employment centres, on the other hand, has a strong negative correlation with the its   
a) number of older buildings   
b) nitrogen oxide levels  
c) proportion of non-retail business acres.  
  
In other words, areas farther away from the centres have younger buildings, less traffic, and fewer retail services. Maybe not that surprising.  

Another self-evident correlation in the data can be seen in the lower right corner of the matrix: Low socioeconomic status of the people living in the area (lstat) has a strong negative correlation with the median value of owner-occupied homes (medv). In other words, poorer people own less-valuable properties and vice versa.  

#####Then we standardize the dataset and print out a summary:  
```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# center and standardize variables
boston_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_scaled)
```
#####Now all our variables follow the same standard and are, thus, comparable to each other.  

#####Let's also convert the data to a data frame format for later use:
```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)
```
##### Now, let's create a categorical variable of the crime rate in the Boston dataset and delete the old crime rate variable

```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# save the scaled crim as scaled_crim
scaled_crim <- boston_scaled$crim

# create a quantile vector of crim and print it
bins <- quantile(scaled_crim)

# create a categorical variable 'crime'
crime <- cut(scaled_crim, breaks = bins, include.lowest = TRUE, label=c("low", "med_low", "med_high", "high"))

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)
```
#####The summary of our dataset will now look like this with our new categorical crime variable as the last column. 
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Print summary of the new dataset
summary(boston_scaled)
```
#####Next, we are going to divide the dataset to train and test sets, so that 80% of the data belongs to the train set:
```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

# save the correct classes from test data
correct_classes <- test$crime
```
#####The summary of our training dataset will now look like this:
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
summary(train)
```
#####And the summary of our test dataset like this:
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
summary(test)
```
#####Then we will fit the linear discriminant analysis on the train set using the categorical crime rate as the target variable and all the other variables as predictor variables:
```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)
lda.fit
```

##### Then we will draw the LDA (bi)plot from the data:
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "black", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, pch = classes, col = classes)
lda.arrows(lda.fit, myscale =2)
```
##### From the plot we can see that the closeness of radial highways strongly correlates with high crime rate.

##### Now, we save the crime categories from the test set and then remove the categorical crime variable from the test dataset:

```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)
```
##### Next, We will predict the classes with the LDA model on the test data:
```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)
```
##### Then we will cross tabulate the results with the crime categories from the test set:
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)
```
##### From the results we can see that our model predicts high crime rate quite well. Then, on the other hand, the model doesn't do that well in other categories. So, in sum, if you want to know, where the bad guys are, our model can help! :)

##### OK, let's now reload the Boston dataset and sandardize it.
```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# load MASS and Boston
library(MASS)
data('Boston')

# Scaling the variables
boston_scaled <- as.data.frame(boston_scaled)

# euclidean distance matrix
dist_eu <- dist(boston_scaled)

# look at the summary of the distances
summary(dist_eu)

```
##### Let's now run a k-means algorithm on the dataset and investigate the optimal number of clusters and rerun the algorithm. After testing different amounts of clusters, I ended up with four, which gave me quite a comprehensible visuals:

```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# k-means clustering
km <- kmeans(dist_eu, centers = 4)

# plot the Boston dataset with clusters
pairs(Boston, col = km$cluster)

```

