# Dimensionality reduction techniques

##### This exercise analyses data from the wrangled *Human* dataset with 155 observations in 8 variables (+ countries as row names).

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# load the wrangled Human data
human = read.table("data/human.csv", sep = ";", header = TRUE)
# Dimensions of the data
dim(human)
```

##### Human is a subset of a dataset at http://hdr.undp.org/en/content/human-development-index-hdi by Tuomo Nieminen.  
##### The original data combines several indicators from most countries in the world and the subset has the following variables:

* "Country" = Country name (as row name)
* "GNI" = Gross National Income per capita
* "LifeExp" = Life expectancy at birth
* "ExpEduYrs" = Expected years of schooling 
* "MMR" = Maternal mortality ratio
* "ABR" = Adolescent birth rate
* "FemaleParl" = Percetange of female representatives in parliament
* "EduFemale" = Proportion of females with at least secondary education
* "LabourFemale" = Proportion of females in the labour force.

The structure of the *human* dataset looks like this:
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Print structure of the data
str(human)
```

And its summary like this:
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Print summary of the data
summary(human)
```




