# Logistic Regression

### Data Wrangling

##### I have created a new dataset 'alcohol-joined' from two datasets of alcohol use habits and background info of Math and Portugese language course students, n=382 (198 females and 184 males) with a mean age of 16.6 years.  

##### Data Files:  
* The original datasets are [student-mat.csv](https://github.com/huttusam/IODS-project/blob/master/data/student-mat.csv) and [student-por.csv](https://github.com/huttusam/IODS-project/blob/master/data/student-por.csv)

* The new dataset is [alcohol-joined.csv](https://github.com/huttusam/IODS-project/blob/master/data/alcohol-joined.csv)  

##### Script:  
* The R script used to create the new dataset (*with comments explaining the process*) is [create_alc.R](https://github.com/huttusam/IODS-project/blob/master/data/create_alc.R)  
  
### Analysis

The new dataset 'alc' has 382 observations and 35 variables. The variables are:

```{r echo=FALSE}
alc = read.table("data/alcohol-joined.csv", sep = ";", header = TRUE)
colnames(alc)
```
####The meaning of each variable is as follows:
* 1 school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira) 
* 2 sex - student's sex (binary: 'F' - female or 'M' - male) 
* 3 age - student's age (numeric: from 15 to 22) 
* 4 address - student's home address type (binary: 'U' - urban or 'R' - rural) 
* 5 famsize - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3) 
* 6 Pstatus - parent's cohabitation status (binary: 'T' - living together or 'A' - apart) 
* 7 Medu - mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2 -  5th to 9th grade, 3 -  secondary education or 4 -  higher education) 
* 8 Fedu - father's education (numeric: 0 - none, 1 - primary education (4th grade), 2 -  5th to 9th grade, 3 -  secondary education or 4 -  higher education) 
* 9 Mjob - mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other') 
* 10 Fjob - father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other') 
* 11 reason - reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other') 
* 12 guardian - student's guardian (nominal: 'mother', 'father' or 'other') 
* 13 traveltime - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour) 
* 14 studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours) 
* 15 failures - number of past class failures (numeric: n if 1<=n<3, else 4) 
* 16 schoolsup - extra educational support (binary: yes or no) 
* 17 famsup - family educational support (binary: yes or no) 
* 18 paid - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no) 
* 19 activities - extra-curricular activities (binary: yes or no) 
* 20 nursery - attended nursery school (binary: yes or no) 
* 21 higher - wants to take higher education (binary: yes or no) 
* 22 internet - Internet access at home (binary: yes or no) 
* 23 romantic - with a romantic relationship (binary: yes or no) 
* 24 famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent) 
* 25 freetime - free time after school (numeric: from 1 - very low to 5 - very high) 
* 26 goout - going out with friends (numeric: from 1 - very low to 5 - very high) 
* 27 Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high) 
* 28 Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high) 
* 29 health - current health status (numeric: from 1 - very bad to 5 - very good) 
* 30 absences - number of school absences (numeric: from 0 to 93)  
  
...and then there are the student's grades that are averages of her Mathematics and Portugese language grades:  
  
* 31 G1 - first period grade (numeric: from 0 to 20) 
* 32 G2 - second period grade (numeric: from 0 to 20) 
* 33 G3 - final grade (numeric: from 0 to 20, output target)  

...the new variables, I have created in the data wrangling are:  
  
* 34 alc_use - the average of workday and weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
* 35 high_use - if alc_use is more than 2, then TRUE else FALSE (binary: TRUE or FALSE)  

####Relationship of alcohol consumption to other variables

Let's look at a graphical summarization of the data:

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# access the tidyverse libraries tidyr, dplyr, ggplot2
library(tidyverse);
# draw a bar plot of each variable
gather(alc) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()

```

Interegsting variables to look at, considering their relation to alcohol consumtion, would be, for example, sex, health, absence from classes and final grades. My hypothesis would be that men use more alcohol than women and that high alcohol consumption leads to lower grades, more abscence and lower health status.  

OK, let's see how the aformentioned four variables correlate with alcohol consumption by cross tabulating them and drawing some plots to illustrate the correlations:

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}

# produce summary statistics by sex and mean alcohol use
alc %>% group_by(sex) %>% summarise(count = n(), mean_alcuse = mean(alc_use)) 

# initialise a plot of sex and alc_use
g1 <- ggplot(alc, aes(x = sex, y = alc_use))

# define the plot as a boxplot and draw it
g1 + geom_boxplot() + ylab("Alcohol Use") + ggtitle("Student alcohol consumption by sex")

# produce summary statistics by high_use and mean final grade
alc %>% group_by(high_use) %>% summarise(mean_grade = mean(G3)) 

# initialize a plot of high_use and G3
g2 <- ggplot(alc, aes(x = high_use, y = G3))

# define the plot as a boxplot and draw it
g2 + geom_boxplot() + ylab("grade") + ggtitle("Student final grades by alcohol consumption")

# produce summary statistics by high_use and mean absences
alc %>% group_by(high_use) %>% summarise(mean_abs = mean(absences)) 

# initialise a plot of high_use and absences
g3 <- ggplot(alc, aes(x = high_use, y = absences))

# define the plot as a boxplot and draw it
g3 + geom_boxplot() + ylab("Absences") + ggtitle("Student absences by alcohol consumption")

# produce summary statistics by high_use and health
alc %>% group_by(high_use) %>% summarise(mean_health = mean(health)) 

# initialise a plot of high_use and healt
g4 <- ggplot(alc, aes(x = high_use, y = health))

# define the plot as a boxplot and draw it
g4 + geom_boxplot() + ylab("Health") + ggtitle("Student health by alcohol consumption")

```  

By looking at the graphical presentations above, we can see that male students report bigger alcohol consumption than their female peers. High alcohol consumption also seems to correlate slightly with lower final grades and more absence from classes. On the other hand, alcohol use does not seem to correlate with health, but we have to keep in mind that the participants of this study are very young (15-22 year olds, mean age 16.6 years) and health problems created by alcohol mostly accumulate over time.



