# Clustering and Classification

This exercise analyses data from the *Boston Dataset*, which is a dataset containing Housing Values of Boston, Massachusettes Suburbias.  

```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# access the MASS package
library(MASS)
# load the data
data("Boston")
```
The data has 506 observations and 14 variables  
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Exploring the dimensions of the data
dim(Boston)
```
and its data structure looks like this:
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Exploring the structure of the data
str(Boston)
```
  
#####The variables stand for:
* crim - per capita crime rate by town.
* zn - proportion of residential land zoned for lots over 25,000 sq.ft.
* indus - proportion of non-retail business acres per town.
* chas - Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).
* nox - nitrogen oxides concentration (parts per 10 million).
* rm - average number of rooms per dwelling.
* age - proportion of owner-occupied units built prior to 1940.
* dis - weighted mean of distances to five Boston employment centres.
* rad - index of accessibility to radial highways.
* tax - full-value property-tax rate per \$10,000.
* ptratio - pupil-teacher ratio by town.
* black - 1000(Bk - 0.63)^2^ where Bk is the proportion of blacks by town.
* lstat - lower status of the population (percent).
* medv - median value of owner-occupied homes in \$1000s.
  
#####The source of the date are  
Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. J. Environ. Economics and Management 5, 81–102.  
Belsley D.A., Kuh, E. and Welsch, R.E. (1980) Regression Diagnostics. Identifying Influential Data and Sources of Collinearity. New York: Wiley.  
  
  
#####Now, let's print a summary of the data:  
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Getting the summary of the data
summary(Boston)
```
#####and visualize the data by printing its correlation matrix  
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
library(corrplot, tidyverse)
# calculate the correlation matrix
cor_matrix<-cor(Boston)
# print the correlation matrix
# corrplot(cor_matrix, type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
# visualize the correlation matrix
corrplot(cor_matrix, method="circle")
```

As we can see from the matrix, some variables have a strong positive (blue circles) or negative correlation (red circles). It appears, for example, that the area's property-tax rate (tax) correaltes heavily with access to radial highways (rad). In ther words, the closer the highways, the bigger the tax.  
  
The area's distance from employment centres, on the other hand, has a strong negative correlation with the its   
a) number of older buildings   
b) nitrogen oxide levels  
c) proportion of non-retail business acres.  
  
In other words, areas farther away from the centres have younger buildings, less traffic, and fewer retail services. Maybe not that surprising.  

Another self-evident correlation in the data can be seen in the lower right corner of the matrix: Low socioeconomic status of the people living in the area (lstat) has a strong negative correlation with the median value of owner-occupied homes (medv). In other words, poorer people own less-valuable properties and vice versa.  


  