# Dimensionality reduction techniques  

This exercise analyses data from the wrangled *Human* dataset with 155 observations in 8 variables.  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# load the wrangled Human data
human = read.table("data/human.csv", sep = ";", header = TRUE)
# Dimensions of the data
dim(human)
```

Human is a subset of a dataset at http://hdr.undp.org/en/content/human-development-index-hdi by Tuomo Nieminen.  
###### The original data combines several indicators from most countries in the world and the subset has the following variables by country:  

* "GNI" = Gross National Income per capita
* "LifeExp" = Life expectancy at birth
* "ExpEduYrs" = Expected years of schooling 
* "MMR" = Maternal mortality ratio
* "ABR" = Adolescent birth rate
* "FemaleParl" = Percetange of female representatives in parliament
* "EduFemale" = Proportion of females with at least secondary education
* "LabourFemale" = Proportion of females in the labour force.

The structure of the *human* dataset looks like this:  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Print structure of the data
str(human)
```

And its summary like this:  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Print summary of the data
summary(human)
```

Now, printing correlation matrix of data and visualizing it further with plots:  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Access libraries
library(ggplot2)
library(GGally)
library(dplyr)
library(corrplot)
library(FactoMineR)
library(tidyr)
# visualize the 'human' variables
ggpairs(human)
# compute the correlation matrix and visualize it with corrplot
cor(human) %>% corrplot()

```
  
**Correlations:**  
  
As we can see from above visualizations, *female education* has a strong positive correlation with expected number of years in education, life expectancy and gross national indome per capita (GNI). There's also a high negative correlation with maternal mortality rate (MMR) and adolescent birth rate (ABR). In sum: Providing women with equal opportunity to education is good for everybody and the society as a whole.  
  
*Expected years in education* has a strong positive correlation with life expectancy and a strong negative correlation with both MMR and ABR.  

*Life expectancy* has a strong positive correlation with GNI and even stronger, but negative, correlation with MMR and ABR.  
  
*GNI* correlates positively with ABR and negatively with MMR.  
  
*MMR* has a high positive correlation with ABR.  

**Distributions:**  
  
It can also be seen from the distributions that female education is very evenly distributed among countries. Proportion of females in the labour force, individual's expected years of education and percentage of females in the parliament follow a somewhat normal distibution. Life expectancy varies between coutries with most countries at the higher end of the scale, but some still lagging behind. Gross national income per capita is still very unevenly distributed with most countries at the low end of the scale. Maternal mortality and adolescent birth rate are fortunately low in most countries, but adolescent pregnancies still seem to be an issue violating the rights of children in many parts of the world.  

Let's perform principal component analysis on the data (unstandardized), show the variability captured by the principal components and draw a biplot:  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# perform principal component analysis (PCA) with the SVD method
pca_human <- prcomp(human)
# print out summaries of the variables
summary(pca_human)
# draw a biplot of the PCA and the original variables
biplot(pca_human, choices = 1:2, cex = c(0.8, 1), col = c("grey70", "blue"))
```

Without standardisation GNI, with its biggest variance, seems to be orthogonal (perpendicular) to other variables. Also, all other variables are on top of eachother and crammed to one side of the matrix and so it is impossible to see any correlations between them.  

So, let's standardize the variables and repeat the analysis:  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# standardize the variables
human_std <- scale(human)
# perform principal component analysis (with the SVD method)
pca2_human <- prcomp(human_std)
# print out summaries of the variables
summary(pca2_human)
# draw a biplot of the principal component representation and the original variables
biplot(pca2_human, choices = 1:2, cex = c(0.8, 1), col = c("grey70", "blue"))

```

From the standardized data, we are able to find the correlating variables and can see which of them are affected by PC1 and PC2. MMR and ABR (the right-pointing arrows) have an effect on overall life expectancy and, further, on expected years of schooling and GNI (the left-pointing arrows). Women's health and rights matter and are important, not only to indivisuals themselves, but to the society as a whole!  

Next, we'll load a new dataset called 'Tea' from the Factominer package and look at the dimensions, structure and summary of the data:  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
data(tea)
# Dimensions of the data
dim(tea)
# Structure of the data
str(tea)
# Summary of the data
summary(tea)
```

The data has answers from 300 individuals to 36 questions about their tea drinking habits. Accoring to information found on the Internet, the researchers have asked 300 individuals how they drink tea (18 questions), what are their product's perception (12 questions) and some personal details (4 questions).  

###### Let's choose some interesting 'Tea' data variables:  
* "Tea" = whether the person drinks black tea, Earl Grey or green tea  
* "How" = whether the person drinks her tea raw, with lemon, with milk or with something else  
* "how" = whether the person uses tea bags, unpackaged tea or both  
* "sugar" = whether the person uses sugar with her tea or not  
* "where" = where does the person buy her tea: tea shop, chain store or both  
* "lunch" = does the person have tea with lunch or not.  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# column names to keep in the dataset
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")
# select the 'keep_columns' to create a new dataset
tea_time <- select(tea, one_of(keep_columns))
# look at the summaries and structure of the data
summary(tea_time)
```

Now, let's visualize the dataset:  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# visualize the dataset
gather(tea_time) %>% ggplot(aes(value)) + geom_bar() + facet_wrap("key", scales = "free") + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

As we can see from the visualization of the 'Tea' dataset, most respondents buy their tea from chain stores as tea bags, prefer Earl Grey over other flavours, don't mix anything else in their tea apart from sugar, and don't drink tea with their lunch.  

Next, we'll apply multiple correspondence analysis (MCA) on these variables:  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)
# summary of the model
summary(mca)
``` 

Looking at the results, we can see that:  
a) Dimensions 1 and 2 explain most of the variance, although non of the dimensions explains the variance significantly more than others.  
b) Dimensions 1 and 2 are strongly linked to place of purchase and packaging method, and Dimension 3 to tea type and how people drink it (use of sugar, mixing milk/lemon/other).  
c) Dimension 1 is most contributed by using unpackaged tea, Dimension 2 by the using both tea bags & unpackaged tea, and Dimension 3 by drinking black tea.  

Finally, let's draw a variable biplot of the analysis:  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# visualize MCA
plot(mca, invisible=c("ind"), habillage = "quali")

```

As we can see from the plot, the packaging method and place of purchase appear to be closely linked: It seems that unpackaged tea is often bought from tea shops and tea bags from chain stores. The plot also shows that unpackaged tea and tea shops are very different categories from the rest, and indicates that green tea is more often bought as unpackaged and from a tea store than other tea types. A difference between everyday tea drinking and those sweet moments with a little bit of feeling of luxury?  

